Bayesian exponential random graph modeling of KWL Burial network

### 2020-05-28 analysis/paper/paper.Rmd
work on the background section

<!--archaeological setting-->
The burial data we analyzed in this paper are collected from the upper component of Kiwulan site, an Iron Age settlement site (AD1350-AD1950) located in northeastern Taiwan which experienced the European colonial impacts in the 17th century and the large wave of Chinese immigrants in the 19th century. The excavation revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with in-situ posts were also excavated. The burials were mostly located in the AD section, which is the largest open area at Kiwulan site that provides continuous stratigraphic sections suitable for comparison. Those burials are oriented in an east-west direction located on the north side of the house structures indicated by postholes and in-situ wooden posts. Previous research indicates the presence of economic inequality between burials, but there is debate about whether these represent an egalitarian or stratified society [@Cheng2008; @Hsieh2012]. @Cheng2008 interprets the unequal distribution of prestige materials, such as the golden beads, between burials, as evidence for hierarchy. However, @Hsieh2012 suggests a relatively egalitarian form from the comparative analysis of frequency and proportion of burial goods. She finds that the rich burials were usually associated with elders, which might indicate achieved, rather than inherited, status. The use of prestige goods in burials might be traced back before the European arrival according to historical records from Spanish and Chinese [@Borao2009; @LiandWu2006). Analysing burial data will give an insight into the social reactions of Kiwulan residents and provide a way to examine and explain the changes in social inequality at Kiwulan before and after European contact.

<!--chronology-->
Burials were assigned to the pre-European period, European and post-European period, and Chinese period for network comparison. Since the sedimentary characteristics show vertical and lateral continuity cross units of AD section, depth is used as the first criterion for burial chronology. The chronology is then evaluated by the burials with light grey glazed jars, also called An-ping jars produced most likely in China and circulated in Southeast Asia from the late 16th to 18th century but are often found at sites with the European occupation in the early 17th century in Taiwan [@Berrocal2018; @Hsieh1995; @Ketel2011]. In addition to the diagnostic porcelain, three radiocarbon dates of charcoal associated with three different burials are used as another evidence to evaluate the chronology [@Chen2007]. Some burials that were heavily disturbed by modern construction are excluded to avoid misinterpreting from incomplete data.

@@

We use burial data collected based on published excavation report, and original excavation records with detailed stratigraphy and context description. Burial data provides information about direction, size, depth, age, gender, classes and quantity of grave goods, and spatial relationship between them. We consider the effects of both exogenous and endogenous factors to generate meaningful explanation of the characteristics of burial network and tendency of burials correlate with each other. To be specific, our purpose is to understand the underlying mechanisms of the network structures and how the relationship influence on the pattern of accumulation of wealth during different foreign contact periods. We focus on connectivity, clusters, centralization, and homophily among those burials that could reflect the social relations between individuals guided by social structures. Previous studies have shown that there are some evidence of social inequality observed from Kiwulan but the degree over time and causes remain unclear. To answer our question for the shifts in social inequality to a more wealth differentiation during and after the arrival of the foreign presences, we hypothesize:

@@

Since the accumulation of imported prestige goods in burials is commonly observed, we assign the burials into four classes, high, medium, low, and none, according to the total number of burial goods as an index of wealth.

@@

We used ergm package in R for the model specification. Every term in an ERGM must have an associated algorithm for computing its value for network [@Morris2008]. The terms we selected for network specification include:

@@

The summary presents the statistics that each θ corresponds to the parameter specified in ERGM previously. In general, positive mean indicates positive correlation, while negative mean indicates negative correlation.

>>>>

<!--archaeological setting-->
The burial data we analyzed are collected from the upper component of Kiwulan site, an Iron Age settlement site (AD1350-AD1950) located in northeastern Taiwan, which experienced the European colonial impacts in the 17th century and the large wave of Chinese immigrants in the 19th century. The excavation revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with in-situ posts were also excavated. The burials were mostly located in the AD section, which is the largest open area at Kiwulan site that provides continuous stratigraphic sections suitable for temporal comparison. Those burials are oriented in an east-west direction located on the north side of the house structures indicated by postholes and `in-situ` wooden posts.

Previous studies on burials suggest a presence of uneven distribution of prestige goods across burials, accompanied with a debate over the differentiation represent an economic inequality supporting an egalitarian society or political inequality indicating a stratified society [@Cheng2008; @Hsieh2012]. @Cheng2008 interprets the unequal distribution of prestige materials, such as the golden beads, between burials, as evidence for hierarchy. However, @Hsieh2012 suggests a relatively egalitarian form from the comparative analysis of frequency and proportion of burial goods. She finds that the rich burials were usually associated with elders, which might indicate achieved, rather than inherited, status. The use of prestige goods in burials might be traced back before the European arrival according to historical records from Spanish and Chinese [@Borao2009; @LiandWu2006). Analysing burial data will give an insight into the social reactions of Kiwulan residents and provide a way to examine and explain the changes in social inequality at Kiwulan before and after European contact.

<!--chronology-->
Burials were primarily assigned to the pre-European period, European and post-European period, and Chinese period for network comparison. Since the sedimentary characteristics show vertical and lateral continuity cross units of AD section, depth is used as the first criterion for burial chronology. The chronology is then evaluated by the burials with light grey glazed jars, also called An-ping jars produced most likely in China and circulated in Southeast Asia from the late 16th to 18th century but are often found at sites with the European occupation in the early 17th century in Taiwan [@Berrocal2018; @Hsieh1995; @Ketel2011]. In addition to the diagnostic porcelain, three radiocarbon dates of charcoal associated with three different burials are used as another evidence to evaluate the chronology [@Chen2007]. Some burials that were heavily disturbed by modern construction are excluded to avoid misinterpreting from incomplete data.

@@

We use burial data collected from published excavation report, and original excavation records with detailed stratigraphy and context description. Given that burials in Chinese phase are far less than previous two phases, we only describe our observations for those burials for comparison. Burial data provides information about direction, size, depth, age, gender, classes and quantity of grave goods, and spatial relationship between them. We consider the effects of both exogenous and endogenous factors to generate meaningful explanation of the characteristics of burial network and tendency of burials correlate with each other. To be specific, our purpose is to understand the underlying mechanisms of the network structures and how the relationship influence on the pattern of accumulation of wealth during different foreign contact periods. We focus on connectivity, clusters, centralization, and homophily among those burials that could reflect the social relations between individuals guided by social structures. Previous studies have shown that there are some evidence of social inequality observed from Kiwulan but how does the degree changes over time and what variables decide the changes remain unclear. To answer the research question for the shifts in social inequality to a more wealth differentiation during and after the arrival of the foreign presences, we hypothesize:

@@

Since the accumulation of imported prestige goods in burials is commonly observed, we assign the burials into four classes, high, medium, and none as an index of wealth. In addition to wealth, we also consider another two node variables, age and gender, serving as indicators for distinguishing ascribed or achieved status in a society [citation].

@@

We used ergm package in R for the model specification. Every term in an ERGM has an associated algorithm for computing its value for network [@Morris2008]. The terms we selected for network specification include:

@@

The network statistics presents that each θ corresponds to the parameter specified in ERGM previously. In general, positive mean indicates positive correlation, while negative mean indicates negative correlation.
